---
title: "PACMOSData"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{PACMOSData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Overview

`PACMOSData` is an ExperimentHub data package providing reference datasets for PACMOS analyses of MESOMICS and lungNENomics cohorts.

The package contains metadata and access utilities for:

* MESOMICS reference datasets
* lungNENomics reference datasets
* PACMOS example and test datasets

These datasets are intended to support reproducible multi-omics analyses using the `PACMOS` framework.

## Installation

```{r, eval=FALSE}
BiocManager::install("PACMOSData")
```

## Load package

```{r, eval = FALSE}
library(PACMOSData)
```

## Available datasets

### MESOMICS datasets

The following datasets are available for MESOMICS analyses:

* `archetypes` : Archetype proportions.
* `D_alt_MOFA` : Genomic Alterations.
* `D_cnv_MOFA` : Copy Number Alterations.
* `D_exprB_MOFA` : Gene expression data.
* `D_loh_MOFA` : Loss of heterozygosity.
* `D_met.bodB_MOFA` : Methylation gene body regions.
* `D_met.enhB_MOFA` : Methylation enhancer regions.
* `D_met.proB_MOFA` : Methylation promoter regions.
* `MESOMICS_latent_factors` : MOFA latent factors.

### lungNENomics datasets

The following datasets are available for lungNENomics analyses:

* `Inputs_MOFA` : MOFA input matrices.
* `lungNEN_LF_K4_with_label` : Archetype proportions with discrete molecular group classifications.
* `lungNEN_LF_K4_without_label` : Archetype proportions without discrete molecular group classifications.
* `lungNEN_LFs` : MOFA latent factors.
* `lungNEN_sample_label` : Sample IDs with molecular groups.

### Test datasets

The package also contains small example datasets used for testing and demonstrating PACMOS workflows:

* `lungNEN_test_expr` : Test expression data for lungNENomics cohort.
* `MESOMICS_test_expr` : Test expression data for MESOMICS cohort.

## Data access

Data resources can be loaded using:

```r
library(PACMOSData)

mesomics_data <- load_pacmos_data("MESOMICS")

lungnen_data <- load_pacmos_data("lungNENomics")

test_data <- load_pacmos_data("test")
```

### Related package

The datasets provided in `PACMOSData` are intended for use with the [PACMOS](https://github.com/IARCbioinfo/PACMOS) 
